Storm surge modelling in the Strait of Georgia

### 2014-12-03 RoughWork/manuscript.tex
Fixed up citations

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too high in the Strait of Juan de Fuca and too low in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There can also be a small contribution to increased sea level due to the thermal expansion of water during warmer years set by the El Nino Southern Oscillation (ENSO) \citep{abeys2011extreme}. %discuss this as a Kelvin wave

@@

Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. A statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm ENSO periods \citep{abeys2011extreme}.

%outline
This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain including a description of the ocean model configuration is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

%A section about how we configured the model for SoG 
We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

@@

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from Fisheries and Oceans Canada (citation?) and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior to the surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction, is presented for this hindcast (Figure \ref{fig:tidesurge}). The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.

>>>>

%overview oF SoG/Salish Sea models
Early work by \citet{crean1988mathematical,crean1988numerical} and \citet{stronach1993update} resulted in a series of models covering the Strait of Juan de Fuca, the Strait of Georgia, and Puget Sound using a finite difference discretization of the equations of motion. A barotropic version of their model was used to reproduce storm surge water level elevations with reasonable accuracy, however, this barotropic version did not include the effects of fresh water entering the system from rivers. Later, a baroclinic version of their model examined residual circulation and fresh water flushing rates \citep{stronach1993update, marinone1996model}. This early work pointed to an important connection between the model's bottom friction parametrization and the location and amplitude of the M$_2$ amphidrome. They found that a local increase in the bottom friction coefficient, typically from 0.002 to 0.004, especially in the area of Haro Strait, was necessary in order to adequately reproduce the observed M$_2$ tidal amplitudes and phase lags in the Strait of Georgia \citep{stronach1993update}. Without local adjustments to the bottom friction their M$_2$ amplitudes were too high in the Strait of Juan de Fuca and too low in the Strait of Georgia.  Other studies over a similar domain, such as the barotropic finite element model of \citet{foreman1995tidal}, have avoided local increases in bottom friction. \citet{foreman2004m} used data assimilation techniques to study M$_2$ tidal dissipation without local changes to the bottom friction in a two-dimensional finite element model that did not include baroclinic effects.

@@

%intro to storm surges
Many coastal communities in the Strait of Georgia are at risk to flooding and property damaged caused by storm surges, a natural hazard arising from the combination of a strong wind storm and high tide. The low atmospheric pressure associated with a storm acts as an inverse barometer elevating the sea level. This effect in combination with strong winds pushing water up against the coast can cause flooding, particularly if the storm occurs during an unusually high tide.  There is also a small contribution to increased sea level due to positive sea surface height anomalies propagating northward from the equator along the eastern Pacific coast during El Nino years \citep{strub2002altimeter}.

@@

Storm surge modelling in the Strait of Georgia dates back to \citet{crean1988mathematical} in a barotropic finite difference model. Their work is highlighted in reviews of storm surges in Canadian waters \citep{murty1995storm, danard2003storm}.  Compared to eastern Canada \citep{gray1984preliminary, bernier2006predicting,bobanovic2006forecasting, bernier2007tide, bernier2010tide}, relatively little has been published on storm surge modelling on the west coast of Canada in recent years. However, a storm surge forecasting program is currently operating over a domain that covers the entire British Columbia coastline \citep{StormBC}. Further, a statistical overview of past surge events in British Columbia waters suggests an increase in extreme water level events during warm El Nino periods \citep{abeys2011extreme}.

This article presents the development of a three-dimensional numerical ocean model for the Strait of Georgia and its accuracy in simulating storm surge hindcasts. This baroclinic model includes the effect of stratification and rivers, however, wetting and drying is not represented. Model evaluation has focused on regions where the near-coast slopes are steep and so wetting and drying effects are believed to be unimportant. A detailed overview of the model configuration and domain including a description of the ocean model configuration is provided in section \ref{sec:config}. This section also describes open and surface boundary conditions as well as choices in grid structure and mixing parametrizations. Next, section \ref{sec:model} presents an evaluation of model performance through comparisons of tidal amplitudes and phases with observations. The model's accuracy in producing storm surge hindcasts is also assessed in section \ref{sec:storm}. Finally, a summary and discussion of future research goals is provided in section \ref{sec:diss}.

@@

We have used the Nucleus for European Modelling of the Ocean (NEMO) framework in its regional configuration \citep{madec2012nemo} to develop an ocean model for the Strait of Georgia and Salish Sea. NEMO is a highly modularized tool used for studying ocean physics, ocean-ice interactions, and the bio-geochemical properties of the ocean. NEMO's ocean core solves the three-dimensional hydrostatic equations of motion for an incompressible fluid under the Boussinesq approximation on a structured computational grid. Although not used in the present work, NEMO's options for grid nesting and bio-geochemical coupling make it a useful tool for studying the complex physics and bio-geochemical interactions within the Strait of Georgia. This work focuses on validating the physical configuration of the Salish Sea model, in particular, determining appropriate forcing and boundary conditions for accurate reproduction of tidal amplitudes and phases as well as storm surge elevations. Future work will include bio-geochemical coupling and data assimilation.

@@

Sea surface height at the mouth of Juan de Fuca was set using values from the Tofino tide gauge \citep{DFOObservations} which is outside of, but close to, the western boundary of the model domain.  A monthly climatology was produced using daily averages from 2000-2010, binning them by month, averaging and setting the yearly mean to zero.  For the storm surge simulations, hourly variations in sea surface height were used.  These values are the Tofino tide gauge values, de-tided and with the zero reset as for the climatology. The storm surge simulations also used the sea surface height anomaly from Port Hardy, forced at the northern boundary in Johnstone Strait. The tidal forcing and sea surface height was applied to the barotropic velocity forcing with the NEMO Flather scheme \citep{flather1994storm, madec2012nemo}. At the west, the baroclinic velocities at the boundary were set equal to the values inside the boundary (zero-gradient boundary conditions).  This scheme is not part of core NEMO 3.4, the version used here.  Zero gradient conditions were chosen because the baroclinic velocity at the mouth of Juan de Fuca is primarily estuarine and thus set by density variations between inside and outside the domain. At the north baroclinic velocities were set to zero.

@@

Model results have been compared with observations by calculating both the total modelled water level and the modelled residual. The modelled residual is calculated as the difference in model output between a simulation with all forcing conditions and a simulation forced only with tides and rivers. Water level observations are taken from the Canadian Tides and Water Level Data Archive \citep{DFOObservations} and are used to calculate the observed residual, defined as the observed water level minus tidal predictions. Tidal predictions are calculated using t\_tide \citep{pawlowicz2002classical}, a MATLAB tool that calculates tidal harmonics from a given time series, in this case, from the year prior to the surge event of interest. Additionally, several simulations with varying forcing conditions such as removing atmospheric forcing or removing the sea surface height anomaly at the open boundaries have been completed.

@@

Next, an assessment of the importance of including the tide-surge interaction, which is important in shallow regions due to nonlinear effects and bottom friction \citep{bernier2007tide}, is presented for this hindcast (Figure \ref{fig:tidesurge}). The most significant differences occur at Point Atkinson where the maximum of the surge-only case is 8.7 cm higher than maximum modelled residual, about 10\% of the total surge amplitude.  In contrast, the tide-surge interaction at Victoria is hardly noticeable. Indeed, a spatial map of the difference between the modelled residual and surge-only difference (not shown here) indicates that the tide-surge interaction is more pronounced after the San Juan and Gulf Islands. The passages between these islands are known for vigorous tidal mixing, which may have a nonlinear effect on the surge propagation. However, the effect is a relatively small overestimate of the surge elevation and no changes to the timing of the surge. So, in operational settings, performing a surge-only simulation and adding the predicted tides \textit{a posteriori} could be more computationally efficient with only a small decrease in the accuracy of the prediction.
